---
title: "mttk interactive walkthrough"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE
)
```

# Welcome

This document is a self-contained, chunk-by-chunk walkthrough of the main
**mttk** workflows. You are running it inside a Binder session with mttk and
all its dependencies already installed.

**How to use this document:**

- Run each chunk in order by clicking the green **Run** button (▶) at the top
  right of the chunk, or press **Ctrl + Shift + Enter** (Windows / Linux) or
  **Cmd + Shift + Enter** (Mac).
- Read the text between chunks — it explains what the code does and how to
  interpret the results.
- Modify the code and re-run to experiment.

---

# 1  Load the package

```{r load}
library(mttk)
```

---

# 2  The dataset

mttk ships a synthetic showcase dataset that mimics a genome-resolved
metatranscriptomics experiment. It contains:

- **16 genes** from **4 genomes** measured across **6 samples**
- **3 control** and **3 oxygen-pulse** incubations, each replicated at 3
  stations
- RNA gene counts, genome-level DNA counts, a genome phylogeny, and
  KO / module / pathway link tables

```{r load-data}
x <- makeShowcaseMTTKExperiment()
x
```

### Sample metadata

```{r sample-meta}
as.data.frame(SummarizedExperiment::colData(x))
```

### Genome metadata

```{r genome-meta}
as.data.frame(genomeData(x))
```

### Phylogenetic tree

```{r tree}
genomeTree(x)
```

Two bacterial and two archaeal genomes form two clades.

### What is stored in the object

```{r structure}
names(geneAssays(x))    # gene-level assays
names(genomeAssays(x))  # genome-level assays
names(links(x))         # hierarchical link tables
```

---

# 3  Gene-level analysis

**Question:** Which individual genes change between the control and oxygen-pulse
treatments?

```{r gene-model}
gene_fit <- fitGeneModel(x, variable = "condition")
```

```{r gene-results}
fitTable(gene_fit)
```

```{r gene-plot, fig.width = 6, fig.height = 4}
plotEffects(gene_fit, alpha = 0.05)
```

Each point is one gene. The x-axis shows the log-scale effect of the
oxygen-pulse treatment. Significant genes (q ≤ 0.05) are highlighted in red
(up) or blue (down).

---

# 4  Genome-level analysis

**Question:** Which genomes shift their overall transcriptional activity in
the oxygen-pulse condition?

```{r genome-model}
genome_fit <- fitGenomeModel(x, variable = "condition")
```

```{r genome-results}
fitTable(genome_fit)
```

```{r genome-volcano, fig.width = 5, fig.height = 4}
plotVolcano(genome_fit)
```

```{r genome-tree, fig.width = 6, fig.height = 4}
plotGenomeEffects(genome_fit, tree = genomeTree(x))
```

Tip colour and size reflect the direction and magnitude of the genome
response. The two bacterial genomes (genome_1 and genome_2) branch together
on the left; the two archaeal genomes (genome_3 and genome_4) on the right.

---

# 5  KO-level analysis

**Question:** Which KEGG Orthologs show a shared response across all genomes?

This workflow fits one negative-binomial mixed model per KO. The genome random
intercept accounts for the fact that genes from the same genome share a common
baseline.

```{r ko-model}
ko_fit <- fitKOMixedModel(x, variable = "condition")
```

```{r ko-results}
fitTable(ko_fit)
```

```{r ko-volcano, fig.width = 5, fig.height = 4}
plotVolcano(ko_fit)
```

```{r ko-effects, fig.width = 5, fig.height = 4}
plotEffects(ko_fit)
```

### Significant KOs only

```{r ko-sig}
significantResults(ko_fit)
```

### Per-KO coefficient table

`coefTable()` gives the full coefficient summary including standard errors:

```{r ko-coef}
coefTable(ko_fit)
```

---

# 6  Module-level meta-analysis

**Question:** Do the KOs assigned to each module show a coherent direction of
response?

Rather than re-aggregating counts, the meta-analysis layer summarises the
already-fitted KO effect estimates. No new model is fitted.

```{r module-meta}
module_fit <- fitModuleMetaAnalysis(x, object = ko_fit)
fitTable(module_fit)
```

```{r module-plot, fig.width = 5, fig.height = 3}
plotEffects(module_fit, n = NULL)
```

---

# 7  Genome-domain comparison

**Question:** Does the KO response differ between Bacteria and Archaea?

`fitKOGroupInteractionModel()` adds a genome-group × condition interaction to
the model so that KO responses can be compared directly between the two
domains.

```{r group-model}
group_fit <- fitKOGroupInteractionModel(
    x,
    variable = "condition",
    group    = "domain"
)
fitTable(group_fit)
```

---

# 8  Clade scan

**Question:** Is there a subtree of related genomes with a distinctly shifted
genome response?

`scanGenomeClades()` tests each internal clade in the phylogeny for a shifted
mean relative to the rest of the tree using a permutation test.

```{r clade-scan}
clade_fit <- scanGenomeClades(genome_fit, x, nPerm = 199)
fitTable(clade_fit)
```

```{r clade-plot, fig.width = 6, fig.height = 4}
plotCladeScan(clade_fit, tree = genomeTree(x))
```

---

# 9  What next?

### Interactive workflow finder

Run this in the R console to get a guided recommendation for your own
biological question:

```{r workflow, eval = FALSE}
findWorkflow()
```

### Adapting the examples to your own data

Replace `x <- makeShowcaseMTTKExperiment()` with your own `MTTKExperiment`
and swap `variable = "condition"` for the column name in `colData(x)` that
holds your treatment variable.

### Further reading

- Full vignette: `vignette("MTTKExperiment-overview", package = "mttk")`
- Function reference: `help(package = "mttk")`
- Source and issue tracker: <https://github.com/alberdilab/mttk>
