---
title: "MTTKExperiment Overview"
author: "MTTK authors"
package: MTTK
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
    %\VignetteIndexEntry{MTTKExperiment Overview}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

MTTK is designed for genome-resolved metatranscriptomics where features are not
just a flat list of genes. Genes belong to genomes, genomes belong to a
community, and genes may also participate in parallel annotation systems such as
KEGG orthologs, modules, pathways, or learned co-expression modules.

The first version of the package focuses on one main container:
`MTTKExperiment`.

`MTTKExperiment` extends
`TreeSummarizedExperiment::TreeSummarizedExperiment`, which means it inherits a
well-established Bioconductor representation for assays, feature metadata,
sample metadata, and optional trees.

# Container design

An `MTTKExperiment` stores information at several levels:

- `assays(x)` stores matrices such as RNA counts, DNA counts, normalized values,
  offsets, fitted values, or residuals.
- `rowData(x)` stores feature-level metadata for genes or transcripts.
- `colData(x)` stores sample-level metadata.
- `genomeData(x)` stores one row per genome.
- `links(x)` stores explicit mapping tables such as gene-to-genome,
  gene-to-annotation, or annotation-to-parent relationships.
- `activeHierarchies(x)` records which hierarchies are currently active in a
  given analysis context.
- `rowTree(x)` and `rowLinks(x)` can store true tree-structured hierarchies when
  a hierarchy is naturally represented as a tree.

The important design choice in this early version is that genome metadata and
explicit mapping tables are stored transparently in `metadata(x)$mttk`. This
keeps the class close to existing Bioconductor infrastructure and makes the
stored information easy to inspect.

# Current package structure

At the moment the package is organized around a small number of source files:

- `R/AllGenerics.R` defines the public accessor generics.
- `R/MTTKExperiment-class.R` defines the S4 class and its validity method.
- `R/MTTKExperiment.R` defines the constructor and display method.
- `R/MTTKExperiment-accessors.R` defines the public accessors and replacement
  methods.
- `R/MTTKExperiment-analysis.R` defines the first aggregation and activity
  helpers for running simple analyses on an `MTTKExperiment`.
- `R/MTTKExperiment-utils.R` contains normalization and validation helpers used
  internally by the package.

This keeps the container layer small and easy to extend before modeling,
network, and aggregation layers are added.

# A small example

The package includes both a ready-to-use example object and a helper function
that constructs the same object from code.

```{r}
library(MTTK)
x <- makeExampleMTTKExperiment()

x
```

The same object is also available as packaged data.

```{r}
data("MTTKExample")
MTTKExample
```

# Working with accessors

The package provides a small set of explicit accessors for the MTTK-specific
parts of the object.

```{r}
genomeData(x)
```

```{r}
names(links(x))
```

```{r}
activeHierarchies(x)
```

```{r}
rnaCounts(x)
```

Assay convenience accessors follow a fixed naming convention. In the current
implementation:

- `rnaCounts(x)` reads or sets the assay named `"rna_counts"`.
- `dnaCounts(x)` reads or sets the assay named `"dna_counts"`.

# Running simple analyses

The first analysis helpers focus on transparent summaries that work directly
with explicit link tables.

Genome-level aggregation:

```{r}
genome_counts <- aggregateToGenome(x)
genome_counts
```

Aggregation through a multi-step functional path:

```{r}
module_counts <- aggregateByLink(
    x,
    path = c("gene_to_ko", "ko_to_module"),
    assays = "rna_counts"
)

module_counts
```

RNA-over-DNA activity summaries:

```{r}
feature_activity <- summarizeActivity(x, by = "feature")
genome_activity <- summarizeActivity(x, by = "genome")

feature_activity
genome_activity
```

These helpers currently return `SummarizedExperiment` objects rather than a new
`MTTKExperiment`, because the aggregated features are no longer genes and the
appropriate hierarchy metadata may differ from the original object.

# What is explicit in the first version

The current package version deliberately keeps link tables explicit and
inspectable. It does not yet impose a rigid schema for every mapping table, and
it does not yet automatically reinterpret arbitrary mapping tables during
subsetting or aggregation. That behavior can be added later once the core link
schemas are stabilized.

This conservative design keeps the first implementation predictable:

- assays live in standard Bioconductor assay storage,
- sample and feature metadata stay in standard Bioconductor slots,
- genome-level metadata and mapping tables remain easy to inspect,
- future modeling and network layers can build on the same container without
  changing the basic object layout.

# Next steps

The next implementation layers are expected to add:

- helper utilities for common link schemas,
- model-fitting interfaces,
- aggregation and summarization helpers,
- network and module-oriented analysis classes.
