---
title: "Using mttk for Hierarchical Metatranscriptomics"
author: "mttk authors"
package: mttk
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
    %\VignetteIndexEntry{Using mttk for Hierarchical Metatranscriptomics}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mttk)
```

# Overview

mttk is designed for genome-resolved metatranscriptomics, where the analysis is
not only about a flat gene-by-sample matrix.

In this setting:

- genes are nested within genomes,
- genome abundance can be measured separately from gene RNA,
- genome relationships can be represented explicitly with a phylogenetic tree,
- genes can also be organized into functional hierarchies such as KO, module,
  and pathway.

This vignette uses a small synthetic coastal oxygen-pulse experiment to show
how the same `MTTKExperiment` can support gene-, genome-, KO-, module-, and
pathway-level questions.

# The showcase dataset

The package ships two example objects:

- `MTTKExample` is the smallest quick-start object and is mainly useful for
  inspection and smoke tests.
- `MTTKShowcase` is a richer object used here to demonstrate the main modeling
  workflows on one compact dataset.

```{r}
data("MTTKShowcase")
x <- MTTKShowcase

x
```

The showcase object represents four genomes sampled across six incubations:
three controls and three oxygen-pulse treatments.

```{r}
genomeData(x)
```

The sample metadata include both categorical and continuous variables that are
useful for illustrating different model types.

```{r}
SummarizedExperiment::colData(x)
```

The container keeps gene-level RNA and genome-level data separate but linked:

```{r}
names(geneAssays(x))
names(genomeAssays(x))
names(links(x))
```

The same object can also store a rooted genome phylogeny for tree-based
follow-up analyses.

```{r}
genomeTree(x)
```

The main assay-level accessors are:

```{r}
rnaGeneCounts(x)[1:4, 1:4]
rnaGenomeCounts(x)
dnaGenomeCounts(x)
```

# Analysis Themes

The easiest way to choose an mttk workflow is to start from the biological
question, then decide whether the signal of interest is:

- a change in individual genes or whole genomes,
- a shared response of a function across genomes,
- a heterogeneous response that differs among genomes or clades,
- or a response pattern that follows the genome phylogeny.

This vignette groups the main workflows that way instead of listing every
question sequentially.

## A companion math vignette

This overview is organized around biological questions and practical choices.
If you also want a gentle explanation of why mttk uses negative-binomial
models, why genome random effects are useful, and what the main model
equations mean, see the companion vignette
`How mttk Models Work`.

## Static decision tree

Use this quick tree when you want to move from a biological question to the
most relevant mttk function.

1. What is your primary unit of interest?
- Gene: use `fitGeneModel()`.
- Genome: go to step 2.
- KO: go to step 3.
- Module or pathway: go to step 4.

2. If the unit is a genome:
- Overall response to a condition or continuous variable: `fitGenomeModel()`.
- Coherence within a genome group: `fitGenomeModel()` then
  `fitGenomeGroupMetaAnalysis()`.
- Global phylogenetic signal: `fitGenomeModel()` then
  `fitGenomePhylogeneticSignal()`.
- Responding subtree: `fitGenomeModel()` then `scanGenomeClades()`.
- Phylogenetically corrected mean response: `fitGenomeModel()` then
  `fitGenomePhylogeneticGLS()`.

3. If the unit is a KO:
- Shared KO response across genomes: `fitKOMixedModel()`.
- KO response differs among genomes: `fitKORandomSlopeModel()`.
- Direct KO contrast between two genome groups: `fitKOGroupInteractionModel()`.
- Coherence within a genome group: `fitKORandomSlopeModel()` then
  `fitKOGenomeGroupMetaAnalysis()`.
- Difference between grouped KO responses:
  `fitKORandomSlopeModel()` then `fitKOGenomeGroupMetaAnalysis()`, then
  `compareKOGenomeGroups()`.
- Global phylogenetic signal: `fitKORandomSlopeModel()` then
  `fitKOPhylogeneticSignal()`.
- Responding subtree: `fitKORandomSlopeModel()` then `scanKOClades()`.
- Phylogenetically corrected mean response: `fitKORandomSlopeModel()` then
  `fitKOPhylogeneticGLS()`.

4. If the unit is a module or pathway:
- Coherent KO-level effects within the set: `fitKOMixedModel()` then
  `fitModuleMetaAnalysis()` or `fitPathwayMetaAnalysis()`.
- Total assigned activity: `fitModuleMixedModel()` or
  `fitPathwayMixedModel()`.
- Response differs among genomes: `fitModuleRandomSlopeModel()` or
  `fitPathwayRandomSlopeModel()`.
- Direct contrast between genome groups: `fitModuleGroupInteractionModel()` or
  `fitPathwayGroupInteractionModel()`.
- Global phylogenetic signal: `fitModuleRandomSlopeModel()` or
  `fitPathwayRandomSlopeModel()`, then `fitModulePhylogeneticSignal()` or
  `fitPathwayPhylogeneticSignal()`.
- Responding subtree: `fitModuleRandomSlopeModel()` or
  `fitPathwayRandomSlopeModel()`, then `scanModuleClades()` or
  `scanPathwayClades()`.
- Phylogenetically corrected mean response: `fitModuleRandomSlopeModel()` or
  `fitPathwayRandomSlopeModel()`, then `fitModulePhylogeneticGLS()` or
  `fitPathwayPhylogeneticGLS()`.

Then refine the call:

- Use `variable = "..."` for a single categorical or continuous predictor.
- Use `formula = ~ ... + ...` and `term = "..."` when you want additional
  covariates but still want one focal effect.
- Add `sampleBlock = "..."` for paired or repeated-measures samples.
- Set `genomeOffset = TRUE` or `FALSE` deliberately when genome abundance
  normalization matters for the biological question.
- Use `genomeCorrelation = "brownian"` where supported when related genomes
  should be correlated in the count model.

If you prefer the same logic in R, use `findWorkflow()`.

## Gene and genome response questions

Use these workflows when the primary response unit is a gene or a whole genome.

- `fitGeneModel()`: "Which transporter, stress-response, or respiration genes
  change between treatments?"
- `fitGenomeModel()`: "Which MAGs shift their overall transcriptional activity
  between conditions, or along a salinity or oxygen gradient?"
- `fitGenomeGroupMetaAnalysis()`: "Do genomes from the same domain or clade
  respond coherently?"

## Function-centered questions

Use these workflows when the main biological unit is a KO, module, or pathway.

- `fitKOMixedModel()`: "Which KOs respond across genomes?"
- `fitModuleMetaAnalysis()` and `fitPathwayMetaAnalysis()`: "Do the KOs
  assigned to this module or pathway show a coherent direction of change?"
- `fitModuleMixedModel()` and `fitPathwayMixedModel()`: "Does the total RNA
  assigned to this higher-level function change?"

## Heterogeneity across genomes or clades

Use these workflows when you expect the same function to behave differently in
different genomes.

- `fitKORandomSlopeModel()`: "Does the same KO increase in some genomes but
  decrease in others?"
- `fitKOGroupInteractionModel()`: "Is that KO response different in Archaea
  versus Bacteria?"
- `fitKOGenomeGroupMetaAnalysis()` and `compareKOGenomeGroups()`: "Within or
  between clades, are KO responses coherent or different?"
- `fitModuleRandomSlopeModel()` and `fitPathwayRandomSlopeModel()`: the same
  question at module or pathway level.

## Phylogeny-aware questions

Use these workflows when the genome tree is biologically more informative than
a single taxonomy column.

- `fitGenomePhylogeneticSignal()`: "Do closely related genomes tend to respond
  similarly overall?"
- `scanGenomeClades()`: "Is there a subtree of genomes with a shifted
  response?"
- `fitKOPhylogeneticSignal()` and `scanKOClades()`: the same questions for KO
  genome-specific responses.
- `fitModulePhylogeneticSignal()` and `fitPathwayPhylogeneticSignal()`: the
  same follow-up for module and pathway random-slope fits.

## Repeated measures or blocked samples

Most modeling workflows also support repeated measures or paired samples
through a sample-level blocking column in `colData(x)`.

In the showcase object, `station_id` links each control incubation to its
matched oxygen-pulse incubation, so the same station can be modeled as a
sample-level random intercept.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitKOMixedModel(
    x,
    variable = "condition",
    sampleBlock = "station_id",
    genomeOffset = FALSE
)
```

Use that option when the question is still gene/KO/genome/module/pathway
response, but the samples are not independent because they come from matched
stations, subjects, vessels, or repeated sampling units.

# Group 1: Expression Shifts In Genes And Genomes

These workflows are most useful when the biological interpretation stays close
to the measured entity itself.

Typical biological questions are:

- "Which stress-response genes increase after an oxygen pulse?"
- "Which genes are associated with dissolved oxygen concentration?"
- "Which MAGs show an overall transcriptional activation under treatment?"
- "Which MAGs track a salinity gradient across incubations?"

## Gene-level changes

For a categorical variable such as treatment group, `fitGeneModel()` gives the
most direct gene-level answer.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
gene_fit <- fitGeneModel(x, variable = "condition")
fitTable(gene_fit, sortBy = "q_value", n = 6)
```

For a continuous variable, fit the same model with the continuous column
directly. In the showcase experiment, `oxygen` can be read as an oxygen
availability gradient across incubations.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitTable(
    fitGeneModel(x, variable = "oxygen"),
    sortBy = "q_value",
    n = 6
)
```

If you need a focal categorical effect while also adjusting for a continuous
covariate, use a formula and select the tested term explicitly.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitGeneModel(
    x,
    formula = ~ condition + oxygen,
    term = "condition"
)
```

## Genome-level shifts

Genome-level modeling is the right choice when the question moves from genes to
MAGs, for example "which genomes become more transcriptionally active during
oxygen exposure?"

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
genome_fit <- fitGenomeModel(x, variable = "condition")
fitTable(genome_fit, sortBy = "q_value")
```

The same interface also supports continuous predictors. For example, this asks
which genomes vary with salinity across incubations.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitTable(
    fitGenomeModel(x, variable = "salinity"),
    sortBy = "q_value"
)
```

If the follow-up question is whether broader genome groups respond coherently,
summarize those genome-level effects by a grouping column in `genomeData(x)`.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitGenomeGroupMetaAnalysis(genome_fit, x, group = "domain")
```

# Group 2: Shared Functional Responses Across Genomes

These workflows are for questions where the main signal of interest is a
function rather than an individual gene.

Typical biological questions are:

- "Which respiratory or oxidative-stress KOs respond across genomes?"
- "Which KOs are associated with low oxygen?"
- "Do the KOs in a module move coherently?"
- "Does total pathway activity increase under treatment?"

## KO-level mixed models

KO-level modeling is the main function-centered workflow in mttk. Gene-level
RNA is first aggregated to KO-within-genome observations, then modeled with a
negative-binomial mixed model.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
ko_fit <- fitKOMixedModel(x, variable = "condition")
fitTable(annotateKOFit(ko_fit, x), sortBy = "q_value", n = 6)
```

This is the right workflow when the main question is categorical, such as
"which KOs change between control and oxygen-pulse incubations?"

The same model also works for continuous variables. Here the question is more
like "which KOs track oxygen availability across samples?"

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
ko_fit_oxygen <- fitKOMixedModel(x, variable = "oxygen")
fitTable(annotateKOFit(ko_fit_oxygen, x), sortBy = "q_value", n = 6)
```

If closely related genomes are expected to share baseline KO activity, replace
independent genome intercepts with a Brownian phylogenetic genome effect.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitKOMixedModel(
    x,
    variable = "condition",
    genomeCorrelation = "brownian"
)
```

## Higher-level functions from KO effects

When the question is whether the member KOs of a module or pathway move in a
coherent direction, keep KO modeling as the primary inferential layer and
summarize upward afterward.

This is often the best match to questions such as:

- "Do oxidative-stress modules show a coordinated response?"
- "Do the KOs in this pathway become more strongly associated with oxygen?"

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
module_meta <- fitModuleMetaAnalysis(ko_fit_oxygen, x, membershipMode = "split")
pathway_meta <- fitPathwayMetaAnalysis(ko_fit_oxygen, x, membershipMode = "split")

fitTable(module_meta, sortBy = "q_value", n = 6)
fitTable(pathway_meta, sortBy = "q_value", n = 6)
```

For overlapping KO annotations, `membershipMode = "split"` is usually the most
conservative default because it divides a KO contribution across its assigned
modules or pathways.

## Total module or pathway activity

Sometimes the biological question is not about coherence of KO effects, but
about the total RNA assigned to a higher-level function. In that case the
appropriate workflow is to re-aggregate counts and refit the model directly at
module or pathway level.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
module_fit <- fitModuleMixedModel(x, variable = "condition")
pathway_fit <- fitPathwayMixedModel(x, variable = "condition")

fitTable(module_fit, sortBy = "q_value", n = 6)
fitTable(pathway_fit, sortBy = "q_value", n = 6)
```

This is the direct count-model answer to questions such as "does total RNA
assigned to this pathway increase during the oxygen pulse?"

There is no count-side `split` mode, because these workflows fit count models,
and splitting memberships would create fractional pseudo-counts instead of
observed counts.

# Group 3: Heterogeneous Responses Across Genomes And Clades

These workflows are for questions where a shared average effect is not enough.

Typical biological questions are:

- "Does the same KO increase in bacteria but decrease in archaea?"
- "Does this pathway respond strongly in only part of the community?"
- "Are KO responses coherent within domains but different between them?"

## KO random-slope models

Some functions are conserved, but their regulation is not. A KO random-slope
model asks whether the KO-level response varies among genomes rather than
assuming one common genome effect.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
ko_rs_fit <- fitKORandomSlopeModel(x, variable = "condition", genomeOffset = FALSE)
fitTable(ko_rs_fit, sortBy = "q_value", n = 6)
```

The genome-specific conditional KO effects can then be inspected directly.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
ko_genome <- koGenomeEffects(ko_rs_fit)
ko_genome[
    ko_genome$ko_id == "K10002",
    c("ko_id", "genome_id", "conditional_effect_estimate")
]
```

If those genome-specific KO responses are expected to be phylogenetically
structured, the same workflow can fit a joint Brownian intercept-slope
covariance across the genome tree.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitKORandomSlopeModel(
    x,
    variable = "condition",
    genomeCorrelation = "brownian",
    genomeOffset = FALSE
)
```

## Direct KO differences between genome groups

If the biological question is explicitly comparative, such as "is this KO more
strongly induced in Archaea than in Bacteria?", fit a direct interaction model.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
ko_group_fit <- fitKOGroupInteractionModel(
    x,
    formula = ~ condition + salinity,
    term = "condition",
    group = "domain",
    genomeOffset = FALSE
)

fitTable(ko_group_fit, sortBy = "q_value", n = 6)
```

This example combines a categorical tested term (`condition`) with a
continuous adjustment variable (`salinity`).

## Summarizing and comparing KO responses by genome group

If the question is not "does the effect differ directly?" but rather "are the
genome-specific KO responses coherent within a group, and how do groups
compare?", summarize the random-slope fit within genome groups.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
ko_domain_fit <- fitKOGenomeGroupMetaAnalysis(ko_rs_fit, x, group = "domain")
fitTable(ko_domain_fit, sortBy = "q_value", n = 6)
```

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
compareKOGenomeGroups(
    ko_domain_fit,
    reference = "Bacteria",
    contrast = "Archaea"
)
```

## Module and pathway heterogeneity

The same kind of heterogeneity can also be asked at module or pathway level.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
module_rs_fit <- fitModuleRandomSlopeModel(x, variable = "condition", genomeOffset = FALSE)
pathway_rs_fit <- fitPathwayRandomSlopeModel(x, variable = "condition", genomeOffset = FALSE)

fitTable(module_rs_fit, sortBy = "q_value", n = 6)
fitTable(pathway_rs_fit, sortBy = "q_value", n = 6)
```

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
moduleGenomeEffects(module_rs_fit)
```

If the question is directly comparative across groups, use the corresponding
group-interaction models.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitModuleGroupInteractionModel(
    x,
    variable = "condition",
    group = "domain",
    genomeOffset = FALSE
)
```

# Group 4: Phylogenetic Signal And Clade-Localized Responses

These workflows are for questions where the genome tree itself is part of the
biology.

Typical biological questions are:

- "Do closely related genomes respond similarly overall?"
- "Is there a responding subtree of genomes?"
- "For a given KO, does the response follow phylogeny?"
- "Is there a clade where a module is consistently induced?"

## Genome-level phylogenetic follow-up

Start from a fitted genome-level model, then ask whether those genome responses
show global phylogenetic structure.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitGenomePhylogeneticSignal(genome_fit, x, nPerm = 99)
```

If the pattern looks local rather than global, scan the tree for clades with
shifted mean responses.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
scanGenomeClades(genome_fit, x, nPerm = 99, minTips = 2)
```

If the question is instead "what is the average genome response after
accounting for phylogenetic covariance?", use the Brownian GLS follow-up.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitGenomePhylogeneticGLS(genome_fit, x)
```

## KO-level phylogenetic follow-up

The same logic applies to genome-specific KO responses from the random-slope
model.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitKOPhylogeneticSignal(ko_rs_fit, x, nPerm = 99, minGenomes = 3)
scanKOClades(ko_rs_fit, x, nPerm = 99, minTips = 2, minGenomes = 3)
fitKOPhylogeneticGLS(ko_rs_fit, x)
```

These are useful for questions such as:

- "Do closely related genomes regulate this KO similarly?"
- "Is there a responding subtree in which this KO is consistently induced?"
- "What is the average KO response after phylogenetic correction?"

## Module and pathway phylogenetic follow-up

The same phylogenetic follow-up can be applied to module and pathway
random-slope fits.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitModulePhylogeneticSignal(module_rs_fit, x, nPerm = 99, minGenomes = 3)
scanModuleClades(module_rs_fit, x, nPerm = 99, minTips = 2, minGenomes = 3)
fitModulePhylogeneticGLS(module_rs_fit, x)
```

The pathway interface is analogous:

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitPathwayPhylogeneticSignal(pathway_rs_fit, x, nPerm = 99, minGenomes = 3)
scanPathwayClades(pathway_rs_fit, x, nPerm = 99, minTips = 2, minGenomes = 3)
fitPathwayPhylogeneticGLS(pathway_rs_fit, x)
```

# Transparent aggregation helpers

mttk also includes explicit aggregation helpers for users who want to inspect
the intermediate summarized data before fitting a model.

```{r}
aggregateToGenome(x)
aggregateToKOGenome(x)
aggregateByLink(x, path = c("gene_to_ko", "ko_to_module"), assays = "rna_gene_counts")
```

These helpers return `SummarizedExperiment` objects, which makes the
intermediate data easy to inspect and reuse in custom workflows.

# Summary

The same `MTTKExperiment` can support several distinct biological questions,
but those questions fall into a small number of practical analysis themes:

- direct shifts in genes or genomes,
- shared functional responses across genomes,
- heterogeneous responses across genomes or clades,
- and phylogeny-aware response patterns.

The key is to pick the workflow that matches the biological question rather
than only the data level. In particular:

- use gene/genome models when you care about the measured entities directly,
- use KO/module/pathway models when you care about function,
- use random-slope or group-interaction models when you expect heterogeneity,
- and use tree-based follow-up when the genome phylogeny is part of the
  hypothesis itself.
