---
title: "MTTKExperiment Overview"
author: "MTTK authors"
package: MTTK
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
    %\VignetteIndexEntry{MTTKExperiment Overview}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

MTTK is designed for genome-resolved metatranscriptomics where features are not
just a flat list of genes. Genes belong to genomes, genomes belong to a
community, and genes may also participate in parallel annotation systems such as
KEGG orthologs, modules, pathways, or learned co-expression modules.

The first version of the package focuses on one main container:
`MTTKExperiment`.

`MTTKExperiment` extends
`TreeSummarizedExperiment::TreeSummarizedExperiment`, which means it inherits a
well-established Bioconductor representation for assays, feature metadata,
sample metadata, and optional trees.

# Container design

An `MTTKExperiment` stores information at several levels:

- `geneExperiment(x)` names the gene-level layer of the object.
- `geneAssays(x)` stores gene-level matrices such as RNA counts, normalized RNA
  values, offsets, fitted values, or residuals.
- `rowData(x)` stores feature-level metadata for genes or transcripts.
  The column `rowData(x)$genome_id` is the authoritative gene-to-genome
  mapping.
- `colData(x)` stores sample-level metadata.
- `genomeExperiment(x)` stores genome-level assays with one row per genome.
- `genomeAssays(x)` stores the assay collection attached to
  `genomeExperiment(x)`.
- `genomeData(x)` stores the genome-level metadata attached to
  `genomeExperiment(x)`.
- `links(x)` stores explicit mapping tables such as gene-to-genome,
  gene-to-annotation, or annotation-to-parent relationships.
- `activeHierarchies(x)` records which hierarchies are currently active in a
  given analysis context.
- `rowTree(x)` and `rowLinks(x)` can store true tree-structured hierarchies when
  a hierarchy is naturally represented as a tree.

The important design choice in this version is that gene-level and genome-level
measurements are stored separately:

- `MTTKExperiment` itself is the gene-level experiment,
- `geneExperiment(x)` and `geneAssays(x)` give that layer an explicit name,
- genome-level assays live in `genomeExperiment(x)`,
- the core gene-to-genome nesting lives in `rowData(x)$genome_id`,
- explicit mapping tables and hierarchy labels remain easy to inspect through
  `links(x)` and `activeHierarchies(x)`.

This makes the common metatranscriptomic layout explicit: genes have RNA
counts, aggregated genome RNA can be stored separately, genomes have DNA
abundances, and genes are linked back to genomes.

# Identifier contract

MTTK now treats the row and column names of the core containers as canonical
identifiers.

- `rownames(x)` are the canonical gene identifiers,
- `colnames(x)` are the canonical sample identifiers,
- `rownames(genomeExperiment(x))` are the canonical genome identifiers.

Metadata columns that mirror those identifiers are optional, but they must
agree exactly when present:

- `rowData(x)$gene_id` must match `rownames(x)`,
- `colData(x)$sample_id` must match `colnames(x)`,
- `genomeData(x)$genome_id` must match `rownames(genomeExperiment(x))`.

This keeps downstream modeling and aggregation code anchored to one
unambiguous set of gene, sample, and genome IDs.

# Link table contract

The first two columns of each table in `links(x)` are treated as the source and
target identifier columns.

- they should use explicit names ending in `_id`,
- a source namespace such as `gene_id` or `genome_id` can resolve directly
  against canonical IDs in the object,
- a downstream source namespace such as `ko_id` or `module_id` must be
  introduced as a target namespace by another link table,
- `gene_to_genome` is the one special case: it is optional and, if stored, it
  must mirror `rowData(x)$genome_id` exactly.

This means that chains such as `gene_to_ko -> ko_to_module ->
module_to_pathway` are valid because each source namespace is anchored either
in the object itself or in an upstream link target.

# Current package structure

At the moment the package is organized around a small number of source files:

- `R/AllGenerics.R` defines the public accessor generics.
- `R/MTTKExperiment-class.R` defines the S4 class and its validity method.
- `R/MTTKExperiment.R` defines the constructor and display method.
- `R/MTTKExperiment-accessors.R` defines the public accessors and replacement
  methods.
- `R/MTTKExperiment-analysis.R` defines the first aggregation and activity
  helpers for running simple analyses on an `MTTKExperiment`.
- `R/MTTKExperiment-utils.R` contains normalization and validation helpers used
  internally by the package.

This keeps the container layer small and easy to extend before modeling,
network, and aggregation layers are added.

# A small example

The package includes both a ready-to-use example object and a helper function
that constructs the same object from code.

```{r}
library(MTTK)
x <- makeExampleMTTKExperiment()

x
```

The same object is also available as packaged data.

```{r}
data("MTTKExample")
MTTKExample
```

# Working with accessors

The package provides a small set of explicit accessors for the MTTK-specific
parts of the object.

```{r}
geneExperiment(x)
```

```{r}
genomeExperiment(x)
```

```{r}
names(geneAssays(x))
```

```{r}
names(genomeAssays(x))
```

```{r}
genomeData(x)
```

```{r}
names(links(x))
```

```{r}
activeHierarchies(x)
```

```{r}
rnaGeneCounts(x)
```

```{r}
dnaGenomeCounts(x)
```

Assay convenience accessors follow a fixed naming convention:

- `rnaGeneCounts(x)` reads or sets the gene-level assay named
  `"rna_gene_counts"`.
- `rnaGenomeCounts(x)` reads or sets the genome-level assay named
  `"rna_genome_counts"`.
- `dnaGenomeCounts(x)` reads or sets the genome-level assay named
  `"dna_genome_counts"`.

# Running simple analyses

The first analysis helpers focus on transparent summaries that work directly
with explicit link tables.

Genome-level aggregation of gene-level RNA. The resulting assay is renamed to
`"rna_genome_counts"` so it is easy to distinguish from stored gene-level RNA.

```{r}
genome_rna <- aggregateToGenome(x)
genome_rna
```

Aggregation through a multi-step functional path:

```{r}
module_counts <- aggregateByLink(
    x,
    path = c("gene_to_ko", "ko_to_module"),
    assays = "rna_gene_counts"
)

module_counts
```

Gene- and genome-level RNA-over-DNA activity summaries:

```{r}
gene_activity <- summarizeActivity(x, by = "gene")
genome_activity <- summarizeActivity(x, by = "genome")

gene_activity
genome_activity
```

These helpers currently return `SummarizedExperiment` objects rather than a new
`MTTKExperiment`, because the aggregated features are no longer genes and the
appropriate hierarchy metadata may differ from the original object.

# KO-level mixed models

The first hierarchy-aware modeling workflow in MTTK fits one mixed model per
KO. Gene-level RNA counts are first aggregated to KO-within-genome
observations, and the model is then fit with `glmmTMB`.

The first implementation supports two model specifications:

- `model = "library_only"` uses
  `rna_count ~ variable + offset(log(lib_size)) + (1 | genome_id)`
- `model = "library_plus_genome_abundance"` adds a genome-abundance offset

The same workflow can be inspected in two explicit preparation steps before the
model is fit:

```{r}
ko_genome <- aggregateToKOGenome(x)
ko_genome
SummarizedExperiment::rowData(ko_genome)[, c("ko_id", "genome_id", "n_genes_pair")]

ko_model_data <- makeKOMixedModelData(x, variable = "condition")
head(as.data.frame(ko_model_data))
S4Vectors::metadata(ko_model_data)$mttk_ko_model$effectLabel
```

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
ko_fit <- fitKOMixedModel(x, variable = "condition")
ko_fit
fitInfo(ko_fit)$effectLabel
as.data.frame(ko_fit)[, c("ko_id", "effect_label", "estimate", "q_value", "status")]
```

The resulting `MTTKFit` can then be ranked, filtered, and annotated:

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitTable(ko_fit, sortBy = "q_value")
significantResults(ko_fit)
annotateKOFit(ko_fit, x)
```

This workflow is aimed at KO-level differential expression or association
analysis, where each KO may be observed in multiple genomes and the genome term
should be modeled explicitly as a random intercept.

# Subsetting behavior

`MTTKExperiment` now defines explicit subsetting behavior for the nested
gene/genome layout.

- row subsetting keeps the selected genes,
- the genome-level companion experiment is pruned to the genomes still
  referenced by `rowData(x)$genome_id`,
- dependent link tables are pruned to identifiers that remain reachable from
  the retained genes,
- column subsetting narrows both the gene-level and genome-level assays to the
  selected samples.

This keeps ordinary `x[i, j]` operations consistent with the biological nesting
stored in the object.

```{r}
x_small <- x[1:2, 1:2]
x_small
names(links(x_small))
genomeData(x_small)
```

# What is explicit in the first version

The current package version deliberately keeps link tables explicit and
inspectable. It does not yet impose a rigid schema for every mapping table, and
it does not yet automatically reinterpret arbitrary mapping tables during
subsetting. The one exception is the core biological nesting of genes within
genomes, which is defined directly by `rowData(x)$genome_id`. A stored
`gene_to_genome` link is therefore optional and, when present, must mirror that
column exactly. Broader link-handling rules can be tightened later once the
remaining link schemas are stabilized.

This conservative design keeps the first implementation predictable:

- the gene-level and genome-level layers are both explicitly named,
- gene-level assays live in the main Bioconductor assay storage,
- genome-level assays live in `genomeExperiment(x)`,
- sample and feature metadata stay in standard Bioconductor slots,
- genome-level metadata and mapping tables remain easy to inspect,
- explicit assay names distinguish `rnaGeneCounts`, `rnaGenomeCounts`, and
  `dnaGenomeCounts`,
- future modeling and network layers can build on the same container without
  changing the basic object layout.

# Next steps

The next implementation layers are expected to add:

- helper utilities for common link schemas,
- model-fitting interfaces,
- aggregation and summarization helpers,
- network and module-oriented analysis classes.
