---
title: "Using MTTK for Hierarchical Metatranscriptomics"
author: "MTTK authors"
package: MTTK
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
    %\VignetteIndexEntry{Using MTTK for Hierarchical Metatranscriptomics}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(MTTK)
```

# Overview

MTTK is designed for genome-resolved metatranscriptomics, where the analysis is
not only about a flat gene-by-sample matrix.

In this setting:

- genes are nested within genomes,
- genome abundance can be measured separately from gene RNA,
- genome relationships can be represented explicitly with a phylogenetic tree,
- genes can also be organized into functional hierarchies such as KO, module,
  and pathway.

This vignette uses a small synthetic coastal oxygen-pulse experiment to show
how the same `MTTKExperiment` can support gene-, genome-, KO-, module-, and
pathway-level questions.

# The showcase dataset

The package ships two example objects:

- `MTTKExample` is the smallest quick-start object and is mainly useful for
  inspection and smoke tests.
- `MTTKShowcase` is a richer object used here to demonstrate the main modeling
  workflows on one compact dataset.

```{r}
data("MTTKShowcase")
x <- MTTKShowcase

x
```

The showcase object represents four genomes sampled across six incubations:
three controls and three oxygen-pulse treatments.

```{r}
genomeData(x)
```

The container keeps gene-level RNA and genome-level data separate but linked:

```{r}
names(geneAssays(x))
names(genomeAssays(x))
names(links(x))
```

The same object can also store a rooted genome phylogeny for tree-based
follow-up analyses.

```{r}
genomeTree(x)
```

The main assay-level accessors are:

```{r}
rnaGeneCounts(x)[1:4, 1:4]
rnaGenomeCounts(x)
dnaGenomeCounts(x)
```

# Choosing a model

The most useful way to approach MTTK is to start from the biological question.

- Use `fitGeneModel()` when the question is "which individual genes change?"
- Use `fitGenomeModel()` when the question is "which genomes shift overall
  transcription?"
- Use `fitGenomePhylogeneticSignal()` when the question is "do closely related
  genomes tend to respond similarly overall?"
- Use `scanGenomeClades()` when the question is "is there a subtree in the
  genome phylogeny with an unusually high or low response?"
- Use `fitGenomeGroupMetaAnalysis()` when the question is "do genomes from the
  same clade or domain respond coherently?"
- Use `fitKOMixedModel()` when the question is "which functions, represented as
  KOs, respond across genomes?"
- Use `fitKORandomSlopeModel()` when the question is "does the same KO behave
  differently in different genomes?"
- Use `fitKOGroupInteractionModel()` when the question is "does the KO
  response differ directly between two genome groups?"
- Use `fitKOPhylogeneticSignal()` when the question is "for a given KO, do the
  genome-specific KO responses follow the phylogeny?"
- Use `scanKOClades()` when the question is "is there a subtree where a KO
  responds in a distinct way?"
- Use `fitKOGenomeGroupMetaAnalysis()` and `compareKOGenomeGroups()` when the
  question is "within or between clades, do KO responses differ?"
- Use `fitModuleMetaAnalysis()` or `fitPathwayMetaAnalysis()` when the question
  is "do the KOs assigned to this higher-level function show a coherent
  response?"
- Use `fitModuleMixedModel()` or `fitPathwayMixedModel()` when the question is
  "does total module/pathway activity change?"
- Use `fitModuleGroupInteractionModel()` or
  `fitPathwayGroupInteractionModel()` when the question is "does the
  module/pathway response differ directly between two genome groups?"
- Use `fitModuleRandomSlopeModel()` or `fitPathwayRandomSlopeModel()` when the
  question is "does the same module/pathway respond differently in different
  genomes?"

The important distinction is between:

- effect-size synthesis upward from KO results, and
- re-aggregating counts and refitting a model at a higher level.

Those answer different questions and should be chosen deliberately.

## Repeated Measures Or Blocked Samples

Most modeling workflows also support repeated measures or paired samples
through a sample-level blocking column in `colData(x)`.

In the showcase object, `station_id` links each control incubation to its
matched oxygen-pulse incubation, so the same station can be modeled as a
sample-level random intercept.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitKOMixedModel(
    x,
    variable = "condition",
    sampleBlock = "station_id",
    genomeOffset = FALSE
)
```

Use that option when the question is still KO/gene/genome/module/pathway
response, but the samples are not independent because they come from matched
sites, subjects, vessels, or repeated sampling units.

# Question 1: Which genes change?

Suppose you want to identify genes whose RNA abundance shifts between control
and oxygen-pulse incubations, while optionally normalizing by parent-genome
abundance.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
gene_fit <- fitGeneModel(x, variable = "condition")
fitTable(gene_fit, sortBy = "q_value", n = 8)
```

This is the most direct workflow when the target of inference is the individual
gene.

If you want to adjust for additional sample covariates while still testing one
focal effect, use a formula and select the term explicitly.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitGeneModel(
    x,
    formula = ~ condition + oxygen,
    term = "condition"
)
```

# Question 2: Which genomes shift overall activity?

Now suppose the question moves from genes to genomes: which MAGs show an
overall transcriptional response after accounting for library size and genome
abundance?

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
genome_fit <- fitGenomeModel(x, variable = "condition")
fitTable(genome_fit, sortBy = "q_value")
```

If the next question is whether broader groups of genomes respond coherently,
the genome-level results can be summarized by a grouping column in
`genomeData(x)`.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
genome_domain_fit <- fitGenomeGroupMetaAnalysis(genome_fit, x, group = "domain")
genome_domain_fit
```

This is useful for questions such as "do archaeal MAGs tend to respond
similarly?" even when the individual genomes are not identical in magnitude.

## Following genome responses with the phylogeny

When the genome relationships are better represented by a tree than by a single
label column, MTTK can test whether genome-level responses show overall
phylogenetic signal.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitGenomePhylogeneticSignal(genome_fit, x, nPerm = 99)
```

That workflow asks whether more closely related genomes tend to have more
similar fitted responses.

If the question is more local than global, scan the rooted tree for clades with
shifted mean responses.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
scanGenomeClades(genome_fit, x, nPerm = 99, minTips = 2)
```

This is useful for questions such as "is there a subtree of related genomes
that becomes more transcriptionally active during the oxygen pulse?"

If the question is instead "what is the average genome response after
accounting for phylogenetic covariance?", use the Brownian GLS follow-up.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitGenomePhylogeneticGLS(genome_fit, x)
```

This uses the fitted genome responses as the input and estimates a
phylogenetically corrected mean response rather than a global signal statistic.

# Question 3: Which KOs respond across genomes?

KO-level modeling is the main function-centered workflow in MTTK. Gene-level
RNA is first aggregated to KO-within-genome observations, then modeled with a
negative-binomial mixed model.

In a real study, this is the right question when you care about conserved
functions such as stress response, respiration, or nitrogen metabolism across
many genomes at once.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
ko_fit <- fitKOMixedModel(x, variable = "condition")
fitTable(annotateKOFit(ko_fit, x), sortBy = "q_value")
```

Here the mixed model uses `genome_id` as a random intercept, so the same KO can
be modeled across multiple parent genomes.

If closely related genomes are expected to share baseline KO activity, use a
Brownian phylogenetic genome effect instead of independent genome intercepts.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitKOMixedModel(x, variable = "condition", genomeCorrelation = "brownian")
```

The same workflow can adjust for additional sample covariates while keeping one
KO-level condition term as the target of inference.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitKOMixedModel(
    x,
    formula = ~ condition + oxygen,
    term = "condition"
)
```

# Question 4: Does the same KO behave differently in different genomes?

Some functions are conserved, but their regulation is not. A KO random-slope
model asks whether the KO-level response varies among genomes rather than
assuming one common genome effect.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
ko_rs_fit <- fitKORandomSlopeModel(x, variable = "condition", genomeOffset = FALSE)
fitTable(ko_rs_fit, sortBy = "q_value")
```

The genome-specific conditional KO effects can then be inspected directly.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
ko_genome <- koGenomeEffects(ko_rs_fit)
ko_genome[ko_genome$ko_id == "K10002", c("ko_id", "genome_id", "conditional_effect_estimate")]
```

That is the right follow-up when the question is something like:
"does the same stress-response KO increase in bacteria but decrease in
archaea?"

If those genome-specific KO responses are expected to be phylogenetically
structured, the same workflow can fit a joint Brownian intercept-slope
covariance across the genome tree.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitKORandomSlopeModel(
    x,
    variable = "condition",
    genomeCorrelation = "brownian",
    genomeOffset = FALSE
)
```

## Directly testing KO differences between genome groups

If the question is specifically whether the KO response differs between two
genome groups, fit a direct interaction model rather than summarizing random
slope estimates afterward.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
ko_group_fit <- fitKOGroupInteractionModel(
    x,
    formula = ~ condition + salinity,
    term = "condition",
    group = "domain",
    genomeOffset = FALSE
)

fitTable(ko_group_fit, sortBy = "q_value")
```

This is useful for questions such as "is the condition effect for this KO
different in Archaea versus Bacteria, after adjusting for salinity?"

## Summarizing KO responses within genome groups

If you want to ask whether genomes in the same domain or clade show a coherent
response for a KO, summarize the KO genome effects by a genome grouping
variable.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
ko_domain_fit <- fitKOGenomeGroupMetaAnalysis(ko_rs_fit, x, group = "domain")
fitTable(ko_domain_fit, sortBy = "q_value")
```

This workflow meta-analyzes the genome-specific KO effects within each group.

## Comparing KO responses between genome groups

If the question is not only whether a KO is coherent within a group, but
whether the KO response differs between groups, compare the grouped KO results
directly.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
compareKOGenomeGroups(
    ko_domain_fit,
    reference = "Bacteria",
    contrast = "Archaea"
)
```

This is especially useful for questions such as "is this KO more strongly
induced in archaea than in bacteria?"

## Following KO responses with the phylogeny

The same KO random-slope fit can also be summarized against the genome
phylogeny instead of a taxonomic label.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitKOPhylogeneticSignal(ko_rs_fit, x, nPerm = 99, minGenomes = 3)
```

This asks whether closely related genomes tend to show similar KO-specific
responses.

If the goal is to localize that pattern, scan the rooted tree for clades with
distinct KO responses.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
scanKOClades(ko_rs_fit, x, nPerm = 99, minTips = 2, minGenomes = 3)
```

This is the right follow-up for questions such as "is there a related subset of
genomes in which this KO is strongly induced, while the rest of the tree shows
a weaker or opposite response?"

If the question is "what is the average KO response after accounting for
phylogenetic covariance among genomes?", fit the Brownian GLS summary instead.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitKOPhylogeneticGLS(ko_rs_fit, x)
```

This is useful when the goal is a phylogenetically corrected KO-level mean
response, rather than a signal test or clade-localization step.

# Question 5: Do higher-level functions show coherent KO-level responses?

When a module or pathway contains several KOs, one useful question is whether
the KO-level responses within that set are moving in a coherent direction.

That is what the module/pathway meta-analysis layer is for.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
module_meta <- fitModuleMetaAnalysis(ko_fit, x)
pathway_meta <- fitPathwayMetaAnalysis(ko_fit, x)

fitTable(module_meta, sortBy = "q_value")
fitTable(pathway_meta, sortBy = "q_value")
```

This is a good choice when you want to keep KO-level modeling as the primary
inferential layer and summarize upward afterward.

If the question is instead whether the higher-level function responds
differently between two genome groups, use a direct interaction model.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
module_group_fit <- fitModuleGroupInteractionModel(
    x,
    variable = "condition",
    group = "domain",
    genomeOffset = FALSE
)

pathway_group_fit <- fitPathwayGroupInteractionModel(
    x,
    variable = "condition",
    group = "domain",
    genomeOffset = FALSE
)

fitTable(module_group_fit, sortBy = "q_value")
fitTable(pathway_group_fit, sortBy = "q_value")
```

This is the direct count-model answer to questions such as "does this pathway
respond more strongly in archaea than in bacteria?"

## Choosing a membership mode for meta-analysis

KO-to-module and KO-to-pathway mappings can be many-to-many, so MTTK lets you
choose how overlapping memberships are handled.

- `duplicate`: every KO contributes fully to every mapped set.
- `split`: a KO contribution is divided across its memberships.
- `exclusive`: only uniquely assigned KOs are kept.

For overlapping functional annotations, `split` is often the most conservative
default when you want to reduce the influence of broadly annotated KOs.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitModuleMetaAnalysis(ko_fit, x, membershipMode = "split")
```

# Question 6: Does total module or pathway activity change?

Sometimes the target is not coherence of KO effects, but total RNA assigned to
a higher-level function. In that case the right workflow is to re-aggregate the
counts and refit the mixed model at the module or pathway level.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
module_fit <- fitModuleMixedModel(x, variable = "condition")
pathway_fit <- fitPathwayMixedModel(x, variable = "condition")

fitTable(module_fit, sortBy = "q_value")
fitTable(pathway_fit, sortBy = "q_value")
```

This answers a different question from `fitModuleMetaAnalysis()` and
`fitPathwayMetaAnalysis()`.

For count reaggregation, MTTK supports:

- `duplicate` for total assigned activity,
- `exclusive` for uniquely assigned activity only.

There is no count-side `split` mode, because these workflows fit count models,
and splitting memberships would create fractional pseudo-counts instead of
observed counts.

If the genomes are phylogenetically structured, the same total-activity model
can use Brownian covariance among genomes directly in the count fit.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitModuleMixedModel(x, variable = "condition", genomeCorrelation = "brownian")
```

# Question 7: Do modules or pathways respond differently across genomes?

The random-slope module/pathway workflows extend the total-activity model to
allow genome-specific responses.

These are useful when the biological question is:
"does the same pathway increase in some genomes but not in others?"

When related genomes are expected to share both baseline activity and response
direction, the same random-slope workflows can use Brownian phylogenetic
intercept and slope terms.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitModuleRandomSlopeModel(
    x,
    variable = "condition",
    genomeCorrelation = "brownian",
    genomeOffset = FALSE
)
```

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
module_rs_fit <- fitModuleRandomSlopeModel(x, variable = "condition", genomeOffset = FALSE)
pathway_rs_fit <- fitPathwayRandomSlopeModel(x, variable = "condition", genomeOffset = FALSE)

fitTable(module_rs_fit, sortBy = "q_value")
fitTable(pathway_rs_fit, sortBy = "q_value")
```

The genome-specific conditional effects can be extracted just as for KO
random-slope models.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
moduleGenomeEffects(module_rs_fit)
```

The same follow-up can also be asked against the genome phylogeny.

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitModulePhylogeneticSignal(module_rs_fit, x, nPerm = 99, minGenomes = 3)
scanModuleClades(module_rs_fit, x, nPerm = 99, minTips = 2, minGenomes = 3)
fitModulePhylogeneticGLS(module_rs_fit, x)
```

These three functions answer different questions:

- `fitModulePhylogeneticSignal()` asks whether closely related genomes tend to
  have similar module responses.
- `scanModuleClades()` asks whether there is a subtree with a shifted module
  response.
- `fitModulePhylogeneticGLS()` asks for the average module response after
  accounting for phylogenetic covariance.

The same interface is available for pathways:

```{r, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fitPathwayPhylogeneticSignal(pathway_rs_fit, x, nPerm = 99, minGenomes = 3)
scanPathwayClades(pathway_rs_fit, x, nPerm = 99, minTips = 2, minGenomes = 3)
fitPathwayPhylogeneticGLS(pathway_rs_fit, x)
```

# Transparent aggregation helpers

MTTK also includes explicit aggregation helpers for users who want to inspect
the intermediate summarized data before fitting a model.

```{r}
aggregateToGenome(x)
aggregateToKOGenome(x)
aggregateByLink(x, path = c("gene_to_ko", "ko_to_module"), assays = "rna_gene_counts")
```

These helpers return `SummarizedExperiment` objects, which makes the
intermediate data easy to inspect and reuse in custom workflows.

# Summary

The same `MTTKExperiment` can support several distinct biological questions:

- which genes change,
- which genomes respond,
- which KOs respond across genomes,
- whether KO responses vary among genomes,
- whether clades differ in KO behavior,
- whether higher-level functions show coherent KO responses,
- whether total module/pathway activity shifts,
- whether higher-level functional responses vary among genomes.
- whether genome, KO, module, or pathway responses follow the genome
  phylogeny globally, locally, or in a phylogenetically corrected mean sense.

The key is to pick the workflow that matches the question:

- use KO/meta-analysis workflows for coherent function-level synthesis,
- use re-aggregated count models for total activity,
- use random-slope models when genome-specific heterogeneity is part of the
  biological signal rather than noise to average away.
